\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-12-02 Manuscript/manuscript.tex
PUB

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the theoretical approach, in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., a symmetry, the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecast with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its features was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism,  by which a forecast is preferred to another one, may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contains the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

Finally, to reduce the computational cost,  one can consider only the pairwise interaction instead of the maximum clique. This is the analogue of what is usually performed in solid-state physics (but not always)  when only the interaction between the first neighbour is considered.  Another example is the  Ising model: here we have a lattice of spins that can be or in $\ket{+}$ or in $\ket{-}$ and their interaction is modelled by\cite{murphy2012machine}:

@@

(this last expression can be reconducted to the multivariate gassian if one consider $\boldsymbol{\Delta}=\boldsymbol{\Sigma}^{-1}$ and $\boldsymbol{\eta}=\boldsymbol{\Delta}\boldsymbol{\mu}$. Given the network, we would now move on how the parameters can be achieved. Lets start from a Markov random field in log-linear form \cite{murphy2012machine}:

@@

Besides the accuracy and other features connected with the confusion matrix, another critical feature is algorithm interpretability. This concept represents how much the algorithm explains its predictions and, in general, its mechanics. In other words, how much the algorithm is not a black box. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g. right to explanation)}, in the author opinion, results provided by such an algorithm cannot be considered scientific since the scientific method also requires to explain and not simply almost correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the formal theory T can be translated into S if and only if S can prove the theorem of T in its language. On the other side, we would also know that an algorithm, as an explanation, is complete: we expect that it will also be able to make correct forecasts for all available data. However, as shown by \cite{doshi2017towards}, and by \cite{gilpin2018explaining} the interpretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal systems as algorithms or scientific theories. For instance, let us consider the fall of object: at first glance, one can consider only the items on Earth. In this case, the acceleration is constant, and a straightforward theory will be obtained.  As one moves to consider also the interaction within planets and stars a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete; the second one is more difficult to explain but much more complete.  This is something that was elegantly stated by A.Einstein \cite{physics-reality}:

@@

Therefore, in principle, the evaluation of algorithm performances should be considered not only into one dimension, the accuracy but also by considering its interpretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more user-friendly mode the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since tit cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, I considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, I constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly selected \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

@@

The first inspection that was performed on the dataset was related to the density distributions of a selection of continuous features that are known, from the celestial mechanics, to be important for the prediction of the asteroids dangerousness. These are reported in Fig. \ref{Density_relevant}. Next we moved on a more systematic analysis with FADM and mutual information. In Fig. \ref{FADM},\ref{FAMD_Quantitative variables} and \ref{FAMD_Individuals_(c)} are reported the main achivements of FADM (performed with FactoMineR package \cite{le2008factominer}): in particular from the correlation circle in Fig. \ref{FAMD_Quantitative variables} the different correlations that are given by the celestial mechanics laws can be recognized: for instance, see that there is strong anti-correlation of the mean motion with the semi-major axis. This is due to the Kepler laws discussed in Appendix A. Furthermore we see that that the mean motion is correctly almost independent with respect to the the diameter (max or min) of  the asteroid. As explained in Appendix A this is an another result of Newton gravitation theory for a two body interaction: the mean motion of an asteroid, as long as the the mass of it is much lower with respect to the sun, is independent from its mass. On the other side the fact that relative velocity is correlated with the diameter is a spurios correlation. At this point one can ask why the mean motion and the relative velocity are orthogonal: this point will be clarified from a theoretical point of view in appendix A and also with graphical models. Briefly the motion of an object on an ellipse as seen from a focus is not uniform: this is faster as the two bodies approaches. Beside this inspection a further analysis based on the concept of mutual information (summarized in the previous chapter) was performed: its result is reported in Fig. \ref{Mutual_information}. This figure summarize the ranking of the features considered in the dataset for the dangerousness classification of the asteroids according to the following expression \cite{kratzer2018varrank}

@@

following the Peng. et al approach \cite{peng2005feature}. It can be seen that the first place in the mutual information ranking, looking to the diagonal element, is taken by minimum orbit intersection: this is correct since this is one parameter used by NASA for deciding if an asteroid is hazardous or not (see Appendix A). The second place is occupied by Epoch date close approach. The third place is the eccentricity value: this parameter is entangled with the minimum orbit intersection and thus it is reasonable that it is important. Then two parameters that are related to the dimension of the asteroid are present: this is meaningful since if the asteroid has a too reduced volume it will be destroyed by the Earth atmosphere. On the other hand the other parameters seems to have a too low MI for being interesting in this preliminary analysis. The results obtained for FAMD and MI will provide a useful path-guide, together with the celestial mechanics theory, for the interpretation of the results obtained in the following section.

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are conditionally dependant on the Hazardous features are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the mean velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) were evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\textwidth]{Figures/ROC_mgm.pdf}
\caption{The ROC curve as obtained from the ROCR package \cite{sing2005rocr} and ggplot2 \cite{ggplot2}. The corresponding $\phi$ value is $0.6$ } 
\label{ROC_mgm}
\end{center}
\end{figure}

@@

\begin{table}[]
\caption{$\phi$ factor for a selected set of ML algorithms as compared with the mgm}
\begin{center}
\begin{tabular}{c|c}
Algorithm & $\phi$ \\ \hline
RF        & 0.9876 \\ \hline
SVM       & 0.7111 \\ \hline
logistic  & 0.6173 \\ \hline
mgm       & 0.5997 \\ \hline
QDA       & 0.5562 
\end{tabular}
\end{center}
\label{phi_values}
\end{table}

@@

\chapter{Conclusions}

The GLASSO (with a penalizing parameter $\rho=0.3$) and the mgm algorithms provided graphical models that correctly capture the physical connections of the dataset feature. In particular, for both models, it can be seen that no one of the orbital parameters is conditionally dependent on the diameter of the asteroid, as stated by theory. Furthermore, the mgm model can provide forecasts about the hazardousness of the asteroids with almost equal or better performances than the logistic and QDA algorithms.  On the other hand, despite the mgm performances being lower than the RF and SVM, its interpretability is higher.  Such a feature compensates the mgm forecast performances gap.

@@

\begin{equation}
(a/a')^{3}\approx(T/T')^{2}
\end{equation}

>>>>

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the theoretical approach, in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., a symmetry, the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecasts with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its features was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as measurable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the ab-initio case, one can fully explain the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism,  by which a forecast is preferred to another one, may not be interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the theory of the involved process is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contains the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. This paper is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset. Finally, the code used to produce the results here reported if provided in Appendix B.

@@

Finally, to reduce the computational cost,  one can consider only the pairwise interaction instead of the maximum clique. This is the analogue of what is usually performed in solid-state physics (but not always)  when only the interaction between the first neighbour atoms is considered.  Another example is the  Ising model: here we have a lattice of spins that can be or in $\ket{+}$ or in $\ket{-}$ and their interaction is modelled by\cite{murphy2012machine}:

@@

(this last expression can be reconducted to the multivariate Gaussian if one consider $\boldsymbol{\Delta}=\boldsymbol{\Sigma}^{-1}$ and $\boldsymbol{\eta}=\boldsymbol{\Delta}\boldsymbol{\mu}$. Given the network, we would now move on how the parameters can be calculated. Lets start from a Markov random field in log-linear form \cite{murphy2012machine}:

@@

Besides the accuracy and other features connected with the confusion matrix, another critical feature is algorithm interpretability. This concept represents how much the algorithm explains its predictions and, in general, its mechanics. In other words, how much the algorithm is not a black box. Apart from all legal problems connected with an algorithm that is a blackbox\footnote{Think for instance the conditions imposed by the GDPR (e.g. right to explanation)}, in the author opinion, results provided by such an algorithm cannot be considered scientific: the scientific method also requires to explain and not simply to provide correct forecast. Following the Tarski et al. \cite{tarski1953undecidable} argument, which is based on mathematical logic, it can be said that the formal theory T can be translated into S if and only if S can prove the theorem of T in its language. On the other side, we would also know that an algorithm, as an explanation, is complete: we expect that it will also be able to make correct forecasts for all available data. However, as shown by \cite{doshi2017towards}, and by \cite{gilpin2018explaining} the interpretability of an algorithm is linked with its incompleteness. This is something that is nested both in formal systems as algorithms or scientific theories. For instance, let us consider the fall of object: at first glance, one can consider only the items on Earth. In this case, the acceleration is constant, and a straightforward theory will be obtained.  As one moves to consider also the interaction within planets and stars, a much more complicate law should be considered (in which the previous case is a particular one). The first one is fully explainable but poorly complete; the second one is more difficult to explain but much more complete.  This is something that was elegantly stated by A.Einstein \cite{physics-reality}:

@@

Therefore the evaluation of algorithm performances should be considered not only into one dimension, the accuracy but also by considering its interpretability as shown in Fig. \ref{ML_intepretability}

@@

The asteroid dataset was retrieved from Kaggle \cite{kaggle_dataset}, which reports into a more machine readable form the dataset of The Center for Near-Earth Object Studies (CNEOS) \cite{cneos+nasa}, a NASA research centre. Among the 40 features that were present in the dataset,  the following cases were excluded: the redundant features (e.g. the distance quantity evaluated in miles instead of kilometres), the features connected only to the other name of the asteroid, the features connected to the name of the orbit and the one connected with the orbiting planet ( since for all it was the Earth). Thus the features obtained were therefore reduced to 22. Their description will be postponed to Appendix A since tit cannot be disjoint with celestial mechanics concepts. Their enumeration is provided in Tab. \ref{tab_features}. Since only 16 \% of the asteroids were hazardous, I considered reducing the number of the non-hazardous to improve the portion of the hazardous ones. For this purpose, I constructed a database in which the proportion hazardous/not hazardous was 1:5: thus, all the hazardous ones were included, while the not-hazardous were randomly extracted \footnote{The author is aware that in principle, the proportion should be much less unbalanced, however, the tool that has to be paid for this operation is an overall reduction of the cases in the dataset that lowers the performances of the algorithms here used. The proportion here is used as a trade-off for these two contrasting facts}. Furthermore, concerning the continuous measures, the dataset was standardised and demeaned.

@@

The first inspection that was performed on the dataset was related to the density distributions of a selection of continuous features that are known, from the celestial mechanics, to be important for the prediction of the asteroids dangerousness. These are reported in Fig. \ref{Density_relevant}. It can be seen that there is an evident distinction for the min orbit intersection, perihelion distance and eccentricity (note that these three parameters are the different faces of the same medal). On the other side, concerning the Absolute magnitude, the distinction seems not so clear. This result, together with the overlap in the three previous features, is because, according to the CNEOS definition (see Appendix A), a hazardous asteroid should have a min orbit intersection lower than a fixed parameter and an absolute magnitude higher to a certain threshold. The fulfilment of both these conditions makes the asteroid dangerous. Next I moved on a more systematic analysis with FADM and mutual information. In Fig. \ref{FADM},\ref{FAMD_Quantitative variables} and \ref{FAMD_Individuals_(c)} are reported the main achivements of FADM (performed with FactoMineR package \cite{le2008factominer}): in particular from the correlation circle in Fig. \ref{FAMD_Quantitative variables} the different correlations that are given by the celestial mechanics laws can be recognized. For instance, see that there is strong anti-correlation of the mean motion with the semi-major axis. This is due to the Kepler laws discussed in Appendix A. Furthermore we see that that the mean motion is correctly almost independent with respect to the the diameter (max or min) of  the asteroid. As explained in Appendix A this is an another result of Newton gravitation theory for a two body interaction: the mean motion of an asteroid, as long as the the mass of it is much lower with respect to the earth, is independent from its mass. On the other side the fact that relative velocity is correlated with the diameter is a spurious correlation. At this point one can ask why the mean motion and the relative velocity are orthogonal: this outcome will be clarified from a theoretical point of view in appendix A and also with graphical models. Briefly the motion of an object on an ellipse as seen from a focus (the Earth) is not uniform: this is faster as the two bodies approaches. Beside this inspection a further analysis based on the concept of mutual information (summarized in the previous chapter) was performed: its result is reported in Fig. \ref{Mutual_information}. This figure summarize the ranking of the features considered in the dataset for the dangerousness classification of the asteroids according to the following expression \cite{kratzer2018varrank}

@@

following the Peng. et al approach \cite{peng2005feature}. It can be seen that the first place in the mutual information ranking, looking to the diagonal element, is taken by minimum orbit intersection: this is correct since this is one parameter used by NASA for deciding if an asteroid is hazardous or not (see Appendix A). The second place is occupied by Epoch date close approach. The third place is the eccentricity value: this parameter is entangled with the minimum orbit intersection and thus it is reasonable that it is important. Then two parameters that are related to the dimension of the asteroid are present: the Estimated min diameter and Absolute Magnitude. This fact is meaningful since if the asteroid has a too reduced volume it will be destroyed by the Earth atmosphere. Indeed the Absolute magnitude and the Min orbit intersection are the features used by CNEOS to classify if an asteroid is hazardous or not (see Appendix A). On the other hand the other parameters seems to have a too low MI score for being interesting in this preliminary analysis.

@@

Let us now move on to the mixed interaction model, where the hazardousness feature was also considered.  The first model was calculated with the mgm package  \cite{mgm,haslbeck2015mgm} where the nodewise regression is used \cite{meinshausen2006high}. The $k$ parameter was set equal to two, and a cross-validation (CV) with ten folds was considered. The result is reported in Fig. \ref{mgm}. From this figure, it can be seen that the features that are conditionally dependant on the Hazardous features are: the absolute magnitude, the min orbit intersection, and the eccentricity.  The theory supports all these connections. The min orbit intersection is the main parameter for the hazardous value. On the other side, the eccentricity can be tough as a parameter that describes how near the celestial body moves at the perihelion (indeed, this parameter is correctly linked with the perihelion distance). The Absolute Magnitude is also a meaningful parameter for the hazard evaluation since, if the asteroid is too small, it will be destroyed by the earth atmosphere.  Furthermore, the parameters related to the diameters are also not connected with the orbital parameters. In addition, the mean velocity is not directly related to the mean motion, but there are the orbital parameters within as expected. Thus it can be said that the model produced is almost consistent with the astronomic laws. On this basis, the performances of the model in terms of confusion matrix, ROC (Receiver operating characteristic) curve and $\phi$ coefficient (also known as Matthews correlation coefficient ) were evaluated (for this purpose the Caret R package \cite{kuhn2008building,caret} was used). The confusion matrix obtained is reported in Fig. \ref{mgm_confusion}: the ROC curve is given in Fig. \ref{ROC_mgm} and it corresponds to a $\phi=0.6$. These performances will be commented on in the next section when the performances of other ML algorithms for the same dataset will be reported. Beside the mgm algorithm/package other approaches were also considered for the evaluation of the mixed interaction graphical model: the function \textit{mmod()} of gRim package \cite{hojsgaard2012graphical} and the \textit{minforest()} of gRapHD (which uses the minForest method) \cite{de2009high}. For both cases, a stepwise algorithm was used.  Their result are reported in Fig. \ref{mmod} and \ref{minforest}. It can be seen that in these models, the minimum diameter is linked with the hazardous feature, which is correct, but what is without a physical meaning is that this quantity is also linked to the features connected with the orbital parameters. A weak link with the orbital parameters is meaningful for the magnitude (as the mgm model for the mean anomaly).  The magnitude depends on the distance between the asteroids and the observer.   However, a direct link from the volume parameters (and thus from the mass) with the orbital ones is not acceptable. In principle, one can set these links as forbidden in the algorithm (blacklist), but in the author view, a model where the result is obtained without boundaries should be preferred to those obtained with a large number of boundaries. Therefore we reject the mmod and minforest ones favouring the mgm, which was obtained without boundaries.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\textwidth]{Figures/ROC_mgm.pdf}
\caption{The ROC (Receiver operating characteristic) curve as obtained from the ROCR package \cite{sing2005rocr} and ggplot2 \cite{ggplot2}. The corresponding $\phi$ value is $0.6$ } 
\label{ROC_mgm}
\end{center}
\end{figure}

@@

\begin{table}[]
\caption{$\phi$ coefficient (also known as Matthews correlation coefficient ) for a selected set of ML algorithms as compared with the mgm}
\begin{center}
\begin{tabular}{c|c}
Algorithm & $\phi$ \\ \hline
RF        & 0.9876 \\ \hline
SVM       & 0.7111 \\ \hline
logistic  & 0.6173 \\ \hline
mgm       & 0.5997 \\ \hline
QDA       & 0.5562 
\end{tabular}
\end{center}
\label{phi_values}
\end{table}

@@

\chapter{Conclusions and Outlook}

The GLASSO (with a penalizing parameter $\rho=0.3$) and the mgm algorithms provided graphical models that correctly reproduce the physical connections of the dataset feature. In particular, for both models, it can be seen that no one of the orbital parameters is conditionally dependent on the diameter of the asteroid, as stated by theory. Furthermore, the mgm model can provide forecasts about the hazardousness of the asteroids with almost equal or better performances than the logistic and QDA algorithms.  On the other hand, despite the mgm performances being lower than the RF and SVM, its interpretability is higher.  Such a feature compensates the mgm forecast performances gap. As a future outlook I consider to extend this analysis for physical phenomena for which there is no deterministic theory but only probabilistic one\footnote{Following the I.Prigogine \cite{prigogine2017non,prigogine1997end,nicolis1989exploring,prigogine1978time} argument it must be said that a pure deterministic theory, if exist, cannot be fully applied. Therefore, in principle, all possible theories should be considered probabilistic. This is because the initial conditions can not be known with a infinite precision. Therefore, if one also perfectly knows the equation of motion and can also solve them exactly (and in most of cases it is not possible) a source of error still persist and it propagates with the Lyapunov exponent. However as it was shown by Prigogine it a system is at equilibrium this instability can neglected. In this case a deterministic approach is meaningful }. For instance one can consider Earthquakes and Vulcanos Eruptions \cite{sornette1989self,sparks2003forecasting,sornette2006critical}

@@

\begin{equation}
\frac{m_{c}+m'}{m_{c}+m}\approx\dfrac{m}{m_{c}}=\left(a/a'\right)^{3}\left( T/T'\right)^{2}
\end{equation}
